---
title: "Forecasting bed occupancy from current admissions"
author: "Thibaut Jombart, Emily S Nightingale, Mark Jit, Olivier le Polain de Waroux, Gwenan M Knight, Stefan Flasche, Rosalind Eggo, Adam J Kucharski, Carl A.B. Pearson, Simon R Procter, CMMID nCov working group & W John Edmunds."
date: "`r format(Sys.time(), '%A %d %B %Y')`"
output:
  html_document:
    code_folding: hide
    highlight: zenburn
    number_sections: yes
    theme: spacelab
    toc: yes
    toc_collapse: no
    toc_depth: 4
    toc_float: yes
    css: !expr here::here('css', 'style.css')
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      collapse = TRUE,
                      fig.width = 8,
                      fig.height = 6,
                      dpi = 150,
                      warning = FALSE,
                      message = FALSE)
```


# Introduction

## What's in this document

This document provides an outline of the model and its implementation. Its code
served as a basis for the app hosted in the same github repository (se `app/`
folder).


## Rationale

This report outlines a model for forecasting bed occupancy from recent data on
admissions. In short, the process is as follows:

1. use a recent number of admissions at a given date as a point of reference

2. simulate growth in admissions from this point of reference using a given
   doubling time, feeding into a log-linear model, including user-specified
   upper and lower bounds
   
3. for each simulated admission, simulate the duration of hospitalisation from
   recent estimate of the duration of hospitalisation; it may be required to
   repeat this process many time for lower numbers of admission, to integrate
   the uncertainty relating to duration of stay
   
4. derive daily numbers of bed needs



## Caveats

* The current model assumes exponential growth

* Default distributions for the duration of stay are provided by matching results of 
[Zhou et al 2020](https://www.thelancet.com/journals/lancet/article/PIIS0140-6736(20)30566-3/fulltext). 
These distributions may need changing under some settings.



## Licensing

* Copyright: Thibaut Jombart
* License: [CC-BY](https://creativecommons.org/licenses/by/4.0/)





<!-- ======================================================= -->
<!-- ======================================================= -->
<!-- ======================================================= -->

# Data preparation {.tabset .tabset-fade .tabset-pills}

## Outline

* **Load scripts**: loads libraries and useful scripts used in the analyses; all
`.R` files contained in `scripts` at the root of the factory are automatically
loaded

* **Load data**: imports datasets, and may contain some *ad hoc* changes to the
data such as specific data cleaning (not used in other reports), new variables
used in the analyses, etc.



## Load packages

```{r libraries}

library(here)
library(reportfactory)
library(incidence)
library(distcrete)
library(epitrix)
library(tidyverse)
library(projections)

```



## Load scripts

These scripts will load:

* all local scripts, stored as `.R` filesinside `/scripts/`
* all global scripts, i.e. stored outside the factory in `../scripts/`

```{r read_scripts}

rfh_load_scripts()

```








<!-- ======================================================= -->
<!-- ======================================================= -->
<!-- ======================================================= -->

# Proof of concept {.tabset .tabset-fade .tabset-pills}

## Model description

We aim to estimate the number of currently circulating cases on a
given day given a number of deaths reported recently. 

The principle of the estimation is:

1. augment the number of admissions using the reporting; currently this is just 
$n_{aug} = n_{reported} / reporting$

2. use a log-linear model, parametrised through the doubling time, to simulate epidemic
   trajectories; see for instance 
   [Jombart et al 2020](https://www.eurosurveillance.org/content/10.2807/1560-7917.ES.2020.25.2.1900735); this is implemented by the RECON package [projections](http://www.repidemicsconsortium.org/projections/)
   
3. for each admission, simulate duration of hospitalisation from provided distribution

4. count beds for each day and simulation


 


## Parameters of the model

This section contain information on the various parameters. We use these
data to generate distribution, with discretisation when needed.

* **Duration of hospitalisation**:
    + non critical care: discretised Weibull(shape:, scale:) to aim for a median of 11
      days, IQR 7-14
    + critical care: discretised Weibull(shape:2, scale:10) to aim for a median of 8
      days, IQR 4-12
    + See table 2 in 
	[source](https://www.thelancet.com/journals/lancet/article/PIIS0140-6736(20)30566-3/fulltext)

```{r los}

## los = "length of stay"
## los_critical for critical care
los_critical <- distcrete("weibull", shape = 2, scale = 10, w = 0, interval = 1)

## los_normal for non critical care hospitalisation
los_normal <- distcrete("weibull", shape = 2, scale = 13, w = 0, interval = 1)

## ## auxiliary function for manual exploration of distributions
## try <- function(shape, scale) {
##   x <- distcrete("weibull", shape = shape, scale = scale, w = 0, interval = 1)$r(1e5)
##   hist(x)
##   summary(x)
## }

```



## Duration of hospitalisation

We visualise the probabiliy mass functions for the first 50 days:

```{r los_plot}

plot(0:50, los_normal$d(0:50),
     type = "h", col = "#e6b800", lwd = 8, lend = 2,
     xlab = "Days in hospital",
     ylab = "Probability",
     main = "Duration of non-critical hospitalisation",
     cex.lab = 1.3, cex.main = 1.5)


plot(0:50, los_critical$d(0:50),
     type = "h", col = "#ff944d", lwd = 8, lend = 2,
     xlab = "Days in hospital",
     ylab = "Probability",
     main = "Duration of critical care hospitalisation",
     cex.lab = 1.3, cex.main = 1.5)

```




## Illustration

As an illustration, we forecast bed occupancy for 2 weeks in a hypothetical location which
would have reported 65 admissions on the current day, with an assumed reporting
of 80% and an assumed R of 2.

```{r illustration}

## input parameters
doubling <- 5 # doubline time in days
doubling_error <- 1.5 # +/- days for upper/lower bounds
reporting <- 0.8
date_start <- Sys.Date() # date of reference
n_start <- 65 # admissions on date_start
duration <- 14 # days of forecasting
last_date <- date_start + duration
los_r <- los_critical$r


## custom graphics
scale_lines <- scale_linetype_manual("", values = c(low = 2, mean = 1, high = 2))

predict_admissions <- function(date_start,
                               n_start,
                               doubling,
                               doubling_error,
                               duration,
                               long = FALSE) {

  ## Sanity checks
  if (!is.finite(n_start)) stop("`n_start` is not a number")
  if (n_start < 1) stop("`n_start` must be >= 1")

  if (!is.finite(doubling)) stop("`doubling` is not a number")
  if (!is.finite(doubling_error)) stop("`doubling_error` is not a number")

  if (!is.finite(duration)) stop("`duration` is not a number")
  if (duration < 1) stop("`duration` must be >= 1")

  ## Outline:

  ## This function calculates future admissions using an exponential model. The
  ## growth rate is calculated from the doubling time, using: r = log(2) / d
  
  ## future dates and initial conditions
  future_dates <- seq(date_start, length.out = duration, by = 1L)
  initial_admissions <- round(n_start / reporting)

  ## calculate growth rate from doubling times
  r <- log(2) / doubling
  r_low <- log(2) / (doubling - doubling_error)
  r_high <- log(2) / (doubling + doubling_error)

  ## calculate future admissions
  future_admissions <- initial_admissions * exp(r * (seq_len(duration) - 1))
  future_admissions_low <- initial_admissions * exp(r_low * (seq_len(duration) - 1))
  future_admissions_high <- initial_admissions * exp(r_high * (seq_len(duration) - 1))
  future_admissions <- round(future_admissions)
  future_admissions_low <- round(future_admissions_low)
  future_admissions_high <- round(future_admissions_high)

  ## build output
  out <- data.frame(date = future_dates,
                    mean = future_admissions,
                    low = future_admissions_low,
                    high = future_admissions_high)

  if (long) {
    out <- tidyr::pivot_longer(out, -1,
                               names_to = "prediction",
                               values_to = "n")
    out$prediction <- factor(out$prediction,
                             levels = c("low", "mean", "high"))
  }
  out
}



## try the function
proj_admissions <- predict_admissions(date_start,
                                      n_start,
                                      doubling,
                                      doubling_error,
                                      duration,
                                      TRUE)

## check outputs and make plots
proj_admissions

proj_admissions %>%
  ggplot(aes(x = date, y = n)) +
  geom_line(aes(linetype = prediction)) +
  scale_lines +
  theme_bw() +
  large_txt +
  rotate_x +
  scale_x_date(date_labels = "%d %b %Y") +
    labs(y = "Daily admissions",
         title = "Forecast of daily admissions")


## simulator for ICU beds
project_beds <- function(dates, n_admissions, r_los, n_sim = 10) {

  ## sanity checks
  if (!length(dates)) stop("`dates` is empty")

  if (!is.finite(n_admissions)) stop("`n_admissions` is not a number")
  if (n_admissions < 1) stop("`n_admissions` must be >= 1")

  if (inherits(r_los, "distcrete")) {
    r_los <- r_los$r
  }
  if (!is.function(r_los)) stop("`r_los` must be a function")

  if (!is.finite(n_sim)) stop("`n_sim` is not a number")
  if (n_sim < 1) stop("`n_sim` must be >= 1")

  
  ## Outline:

  ## We take a vector of dates and incidence of admissions, and turn this into a
  ## vector of admission dates, whose length is sum(n_admissions). We will
  ## simulate for each date of admission a duration of stay, and a corresponding
  ## vector of dates at which this case occupies a bed. Used beds are then
  ## counted (summing up all cases) for each day. To account for stochasticity
  ## in duration of stay, this process can be replicated `n_sim` times,
  ## resulting in `n_sim` predictions of bed needs over time.

  
  admission_dates <- rep(dates, n_admissions)
  n <- length(admission_dates)
  last_date <- max(dates)
  out <- vector(n_sim, mode = "list")
  

  for (j in seq_len(n_sim)) {
    los <- r_los(n)
    list_dates_beds <- lapply(seq_len(n),
                              function(i) seq(admission_dates[i],
                                              length.out = los[i],
                                              by = 1L))
    ## Note: unlist() doesn't work with Date objects
    dates_beds <- do.call(c, list_dates_beds)
    beds_days <- incidence::incidence(dates_beds)
    if (!is.null(last_date)) {
      to_keep <- incidence::get_dates(beds_days) <= last_date
      beds_days <- beds_days[to_keep, ]
    }

    out[[j]] <- projections::build_projections(
                                 x = beds_days$counts,
                                 dates = incidence::get_dates(beds_days))
  }

  projections::merge_projections(out)
 
}



## split predicted admissions by type of prediction (low / mean / high)
list_proj_admissions <- split(proj_admissions, proj_admissions$prediction)



## simulate beds for each epicurve
list_proj_beds <- lapply(list_proj_admissions,
                         function(e)
                           project_beds(e$date,
                                        e$n,
                                        los_normal$r))


## merge results
beds <- merge_projections(list_proj_beds)

## plot results
plot(beds,
     quantiles = c(.025, .5),
     ribbon = TRUE) +
  theme_bw() +
  large_txt +
  scale_x_date(date_label = "%d %b %y") +
  rotate_x +
  labs(title = "Predicted bed occupancy",
       x = NULL,
       y = "Daily numbers of beds")

```








<!-- ======================================================= --> 
<!-- ======================================================= --> 
<!-- ======================================================= -->

# System information {.tabset .tabset-fade .tabset-pills}

## Outline

The following information documents the system on which the document was
compiled.


## System 

This provides information on the operating system.

```{r system_info}
Sys.info()
```

## R environment

This provides information on the version of R used:

```{r R_session}
R.version
```


## R packages

This provides information on the packages used:

```{r R_pkg}
sessionInfo()
```
